Supporting Information : The importance of a complete description of dependence between variables for ecology and related fields

### 2019-05-03 SupportingInformation.Rmd
edits

<!--Display model selection results for Cedar-creek data for year 2000 in a table-->
```{r tab_cc2000fit, echo=F, results='markup'}
h<-BivMS_res_cc$`2000`$InfCritRes[,c("copname","AICw","BICw","LTdep","UTdep")]
colnames(h)<-c("copname","AICw","BICw","LT","UT")
knitr::kable(h,
             format='latex',linesep = "",
             caption.short = "Model fitting results for the Cedar Creek dataset",
             caption = "Results of fitting our 16 copula families to the Cedar Creek dataset. Analogous to Fig. \\ref{tab_soilCNfit}, see caption of that figure for details. \\label{tab_cc2000fit}",
             booktabs=T)
```

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/Ceder_creek_results/2000cc_numbin_4_bivfunctionplot.pdf}
\caption[Nonparametric tests against an N-copula null, Cedar Creek data]{Nonparametric tests for tail dependence and other deviations from N-copula structure, for Cedar Creek data. As described in the main text, the values of the statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$ and $\Dsq_{l_b,u_b}$ for real data (red crosses) were compared to distributions of their values on 1000 Kendall- or Spearman-preserving N surrogates of the data (blue and green x's show 0.025 and 0.975 quantiles), separately in two comparisons for each of the ranges $(l_b,u_b)=(0,0.25),\ldots,(0.75,1)$. If the red cross was outside the range given by the x's, text at the top of panels indicates the number of surrogate values the real-data value was greater than or less than. For instance, a value $>N$ (respectively, $<N$) means the value of the statistic on real data was greater than (respectively, less than) its value on $N$ surrogates. When the statistic was greater than 975 or less than 975 surrogate values, it indicates significance ($95\%$ confidence level). When $\cor$ or $\Ps$ values (respectively, $\Dsq$ values) were greater than surrogates, it means dependence in that part of the distributions was stronger than (respectively, weaker than) expected from a normal-copula null hypothesis.\label{fig_cc2000_nonparam}}
\end{center}
\end{figure}

<!--Displaying a table with npa summary results for Cedar creek data-->
```{r tab_cc_asym_2000, echo=F, results="markup"}
tab_cc_asym_2000<-make_table_stat_fn(data=stat_cc_ga$`2000`)
#tab_cc_asym_2000[,1]<-c("corFS-corLS","PFS-PLS","D2LS-D2FS")
knitr::kable(tab_cc_asym_2000,
             format="latex",linesep = "",escape=F,
             caption.short = "Nonparametric asymmetry tests against an N-copula null, Cedar Creek data",
             caption = "Cedar Creek dataset results for nonparametric tests of whether asymmetry of tail dependence was significant, compared to an N-copula null hypothesis. The values of the listed statistics (cor$_{FS}$-cor$_{LS}$ = $\\cor_{0,0.25}-\\cor_{0.75,1}$, P$_{FS}$-P$_{LS}$ = $\\Ps_{0,0.25}-\\Ps_{0.75,1}$, D$_{LS}^2$-D$_{FS}^2$ = $\\Dsq_{0.75,1}-\\Dsq_{0,0.25}$; here FS stands for `first slice' and LS for `last slice') for real data were compared to their values for 1000 Kendall- or Spearman-preserving N surrogates, in seperate comparisons. A table entry $>X$ indicates the value of the given statistic on the data was higher than its value on $X$ of the surrogates, so entries of the form $>X$ for $X$ equal to $975$ or above would indicate that lower-tail dependence was significantly stronger than upper-tail dependence.\\label{tab_cc_asym_2000}",
             booktabs=T)
```

@@

For bivariate datasets, we compared values 
of the statistics $\cor_{l_b,u_b}$,
$\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ to 
distributions of values of the same statistics computed on surrogate datasets that were produced from 
the empirical data by randomizing it in a special way to have no 
tail dependence (Appendix \ref{surrog_test}). 
The surrogate/randomized datasets had exactly 
the same marginal distributions as the empirical data and had very similar Kendall or Spearman 
correlation (the sorrogate algorithm had two versions, one for preserving each correlation coefficient), 
but had normal copula structure. Thus our comparisons tested the null hypothesis 
that our statistics took values on the empirical data no different from what would have been expected 
if the copula structure of the data were normal, but the data were otherwise 
statistically unchanged. The comparison of one of the
statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, as computed on the 
empirical data, to the distribution
of its values computed on many surrogate datasets 
provides a test of whether dependence in the part of the distribution specified 
by $l_b$ and $u_b$ is different from
what would be expected from a null hypothesis of normal copula structure but 
unchanged marginals and similar 
Kendall (or Spearman)
correlation. Thus the comparison helps answer the first two parts of Q1
for bivariate datasets. 
Significant deviations correspond to deviations from normal 
copula structure. In particular, deviations 
using $l_b=0$ and $u_b$ small (say, $0.1$) correspond to lower tail dependence 
different from that of a normal copula;
likewise, using $l_b=0.9$ and $u_b=1$ tests for upper tail dependence different 
from that of a normal copula.
The comparison of one of the statistics 
$\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$,
as computed on the empirical data, to the distribution of its values on many surrogate datasets 
provides a test of asymmetry of dependence. Thus this comparison helps answer the
third part of Q1 for bivariate 
datasets.

@@

For first three bivariate datasets in Table \ref{MT-tab_bivar_summary},
the statistic $P_{0,0.1}-P_{0.9,1}$ was computed for real data and 
for 1000 Kendall-preserving surrogate datasets with N copulas. Due to relatively smaller sample size,
for Cedar creek data we used a wider range of boundaries $(l_b,u_b)$ to compute $P_{0,0.25}-P_{0.75,1}$.
For each data, rank of the said statistics with real-data value in 
the distribution of surrogate values was computed. 
If ranks were $>975$ or $<25$, asymmetry of tail 
dependence was significant compared to a N-copula null hypothesis. Ranks were `r unname(tab_bivar_summary[10,])`
for, respectively, the soil C and N data, bird masses and metabolic rates, mammal masses
and metabolic rates, and the Cedar Creek data. Thus the first three datasets showed stronger 
upper- than lower-tail dependence, and the Cedar Creek data which showed the reverse though not significantly with
non-parametric analysis, confirming the 
results based on model selection presented in the main text (see Table \ref{MT-tab_bivar_summary}, row 9 
showed Cedar creek had relatively weak tail-dependence than the other three datasets).

>>>>

<!--Display model selection results for Cedar-creek data for year 2000 in a table-->
```{r tab_cc2000fit, echo=F, results='markup'}
h<-BivMS_res_cc$`2000`$InfCritRes[,c("copname","AICw","BICw","LTdep","UTdep")]
colnames(h)<-c("copname","AICw","BICw","LT","UT")
knitr::kable(h,
             format='latex',linesep = "",
             caption.short = "Model fitting results for the Cedar Creek dataset",
             caption = "Results of fitting our 16 copula families to the Cedar Creek dataset, year 2000. Analogous to Fig. \\ref{tab_soilCNfit}, see caption of that figure for details. \\label{tab_cc2000fit}",
             booktabs=T)
```

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=16 cm]{./Results/Ceder_creek_results/2000cc_numbin_4_bivfunctionplot.pdf}
\caption[Nonparametric tests against an N-copula null, Cedar Creek data]{Nonparametric tests for tail dependence and other deviations from N-copula structure, for Cedar Creek data, year 2000. As described in the main text, the values of the statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$ and $\Dsq_{l_b,u_b}$ for real data (red crosses) were compared to distributions of their values on 1000 Kendall- or Spearman-preserving N surrogates of the data (blue and green x's show 0.025 and 0.975 quantiles), separately in two comparisons for each of the ranges $(l_b,u_b)=(0,0.25),\ldots,(0.75,1)$. If the red cross was outside the range given by the x's, text at the top of panels indicates the number of surrogate values the real-data value was greater than or less than. For instance, a value $>N$ (respectively, $<N$) means the value of the statistic on real data was greater than (respectively, less than) its value on $N$ surrogates. When the statistic was greater than 975 or less than 975 surrogate values, it indicates significance ($95\%$ confidence level). When $\cor$ or $\Ps$ values (respectively, $\Dsq$ values) were greater than surrogates, it means dependence in that part of the distributions was stronger than (respectively, weaker than) expected from a normal-copula null hypothesis.\label{fig_cc2000_nonparam}}
\end{center}
\end{figure}

<!--Displaying a table with npa summary results for Cedar creek data-->
```{r tab_cc_asym_2000, echo=F, results="markup"}
tab_cc_asym_2000<-make_table_stat_fn(data=stat_cc_ga$`2000`)
#tab_cc_asym_2000[,1]<-c("corFS-corLS","PFS-PLS","D2LS-D2FS")
knitr::kable(tab_cc_asym_2000,
             format="latex",linesep = "",escape=F,
             caption.short = "Nonparametric asymmetry tests against an N-copula null, Cedar Creek data",
             caption = "Cedar Creek dataset results for nonparametric tests of whether asymmetry of tail dependence was significant, compared to an N-copula null hypothesis, year 2000. The values of the listed statistics (cor$_{FS}$-cor$_{LS}$ = $\\cor_{0,0.25}-\\cor_{0.75,1}$, P$_{FS}$-P$_{LS}$ = $\\Ps_{0,0.25}-\\Ps_{0.75,1}$, D$_{LS}^2$-D$_{FS}^2$ = $\\Dsq_{0.75,1}-\\Dsq_{0,0.25}$; here FS stands for `first slice' and LS for `last slice') for real data were compared to their values for 1000 Kendall- or Spearman-preserving N surrogates, in seperate comparisons. A table entry $>X$ indicates the value of the given statistic on the data was higher than its value on $X$ of the surrogates, so entries of the form $>X$ for $X$ equal to $975$ or above would indicate that lower-tail dependence was significantly stronger than upper-tail dependence.\\label{tab_cc_asym_2000}",
             booktabs=T)
```

@@

For bivariate datasets, we compared values 
of the statistics $\cor_{l_b,u_b}$,
$\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, $\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$ to 
distributions of values of the same statistics computed on surrogate datasets that were produced from 
the empirical data by randomizing it in a special way to have no 
tail dependence (Appendix \ref{surrog_test}). 
The surrogate/randomized datasets had exactly 
the same marginal distributions as the empirical data and had very similar Kendall or Spearman 
correlation (the surrogate algorithm had two versions, one for preserving each correlation coefficient), 
but had normal copula structure. Thus our comparisons tested the null hypothesis 
that our statistics took values on the empirical data no different from what would have been expected 
if the copula structure of the data were normal, but the data were otherwise 
statistically unchanged. The comparison of one of the
statistics $\cor_{l_b,u_b}$, $\Ps_{l_b,u_b}$, $\Dsq_{l_b,u_b}$, as computed on the 
empirical data, to the distribution
of its values computed on many surrogate datasets 
provides a test of whether dependence in the part of the distribution specified 
by $l_b$ and $u_b$ is different from
what would be expected from a null hypothesis of normal copula structure but 
unchanged marginals and similar 
Kendall (or Spearman)
correlation. Thus the comparison helps answer the first two parts of Q1
for bivariate datasets. 
Significant deviations correspond to deviations from normal 
copula structure. In particular, deviations 
using $l_b=0$ and $u_b$ small (say, $0.1$) correspond to lower tail dependence 
different from that of a normal copula;
likewise, using $l_b=0.9$ and $u_b=1$ tests for upper tail dependence different 
from that of a normal copula.
The comparison of one of the statistics 
$\cor_{0,u_b}-\cor_{1-u_b,1}$, 
$\Ps_{0,u_b}-\Ps_{1-u_b,1}$ and
$\Dsq_{1-u_b,1}-\Dsq_{0,u_b}$,
as computed on the empirical data, to the distribution of its values on many surrogate datasets 
provides a test of asymmetry of dependence. Thus this comparison helps answer the
third part of Q1 for bivariate 
datasets.

@@

For first three bivariate datasets in Table \ref{MT-tab_bivar_summary},
the statistic $P_{0,0.1}-P_{0.9,1}$ was computed for real data and 
for 1000 Kendall-preserving surrogate datasets with N copulas. Due to relatively smaller sample size,
for Cedar Creek data we used $P_{0,0.25}-P_{0.75,1}$.
For each dataset, the rank of the statistic for real-data in 
the distribution of surrogate values was computed. 
If ranks were $>975$ or $<25$, asymmetry of tail 
dependence was significant compared to an N-copula null hypothesis. Ranks were `r unname(tab_bivar_summary[10,])`
for, respectively, the soil C and N dataset, the bird body masses and metabolic rates dataset, 
the mammal body masses and metabolic rates dataset, and the Cedar Creek data for the year 2000. 
Thus the first three datasets showed stronger 
upper- than lower-tail dependence, and the Cedar Creek data showed the reverse 
(though not significantly), confirming the 
results based on model selection presented in the main text (see 
Table \ref{MT-tab_bivar_summary}, row 9). Model selection results for Cedar Creek showed 
only slightly stronger lower- than upper-tail dependence, consistent with the non-significant 
tendency toward lower-tail dependence also revealed by the non-parametric analysis for that
dataset.
